State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-16 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
Changes to section 3 on spectral analysis and model selection

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite-dimensional state-space model. Spatial frequency analysis methods are introduced for model selection by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic dynamics with the proposed framework. The results also illustrate the loss of high spatial frequency information as the cost incurred by the model reduction procedure. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

Generating physiologically plausible neural field models are of great importance for studying brain dynamics at the meso/macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's meso and macro-scale dynamics remains largely theoretical. Understanding the brain at this level is extremely important since it is at this scale that pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested.

@@

Unfortunately, the use of these models in the clinic has been limited, since they are constructed for ``general'' brain dynamics whereas pathologies almost always have unique underlying patient-specific causes. Patient-specific data from EEG is readily available in the clinical setting, suggesting an opportunity to make the patient-specific link to models of cortical dynamics. Furthermore, recent technological advances have driven an increased level of sophistication in recording techniques, with dramatic increases in spatial and temporal sampling~\cite{Brinkmann2009}. However, the meso/macroscopic neural dynamic state is not directly observable in EEG data, making predictions of the underlying physiology inherently difficult.

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the Wilson and Cohen (WC)~\cite{Wilson1973} or Amari~\cite{Amari1977} models or a neural mass model like the Jansen and Ritt model~\cite{Jansen1995}, to patient-specific EEG data. Fitting the neural models to individuals is a highly non-trivial task and, until very recently, has not been reported in the literature.

@@

The paper proceeds by first deriving the continuum neural field equations in Section~\ref{NeuralModelSection}. Then a finite dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. Section~\ref{SpectralAnalysisSection} establishes conditions using spatial frequency analysis for both sensor and basis function spacing and width, such that the dominant dynamics of the neural field can be represented by the reduced model. The state and parameter estimation procedure is described in Section~\ref{StateAndParameterEstimationSection}. The results for the spatial frequency analysis and parameter estimation are then presented in Section~\ref{ResultsSection}. The implications and limitations of this framework are discussed in Section~\ref{DiscussionSection} along with planned future developments.

@@

Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular as they are parsimonious yet have a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009,Qubbaj2007}. In certain regions of the cortex, each column is thought to be connected locally via symmetric short range local excitation with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations and epileptic behaviour~\cite{David2003,Bressloff2001,Breakspear2006}. Field models are also capable of generating complex spatial patterns of activity such as Turing patterns, spirals and travelling oscillations~\cite{Amari1977,Coombes2005,Coombes2007}.

@@

The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at position $\mathbf{r}$ to the local post-synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations at locations $\mathbf{r}'$ can be described by 

@@

where $\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Non-local interactions between cortical populations are described by 

@@

where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel and $\Omega$ is the spatial domain representing a cortical sheet or surface. The connectivity kernel is typically a ``Mexican hat'' function, which describes strong local activation, weak mid-range repression and weak long-range activation. An example of the connectivity kernel  used in this paper can be seen in Figure~\ref{fig:2d_kernel}. The exact shape of this kernel is assumed to vary across patients, and hence needs to be inferred from data.

@@

   	\caption{A Mexican-hat connectivity kernel, $w(\mathbf{r},\mathbf{r'})$, used in this paper. The function $w(.)$ is rotationally symmetric (isotropic) about zero, hence a cross-section captures the important aspect of the connectivity kernel's shape. The kernel decays asymptotically to zero.} 

@@

To arrive at the final form of the model, we shall express the synaptic response kernel as a Green's function 

@@

To arrive at the integro-difference equation (IDE) form of the model, we discretise time using a first-order Euler method (see~\ref{Time Discretization}) giving 

@@

The mapping between the membrane voltage and the collected iEEG/LFP data is modelled using the observation function 
\begin{equation}
    \label{eq:ObservationEquation}
	\mathbf{y}_t =
	\int_{\Omega}{
	    m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)
	d\mathbf{r}'} + 
	\boldsymbol{\varepsilon}_t, 
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field, $n=1,...,N$ indexes the sensors and $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$, $\mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ denotes the multivariate normal distribution with mean zero and covariance matrix $\Sigma_{\varepsilon}$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the sensor pick-up geometry where 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}. 
\end{equation}

@@

Decomposition allows a continuous field to be represented by a finite-dimensional state vector. This facilitates application of standard nonlinear state estimation methods such as the unscented Kalman filter. The field decomposition is described by 

@@

where $\boldsymbol{\psi}(r,r')$ is a vector of Gaussian basis functions and $\boldsymbol{\theta}$ is a vector of scaling parameters. By assuming a Gaussian isotropic connectivity structure, the kernel basis functions can be written as $\psi(\mathbf{r}-\mathbf{r}')$. We will assume that we know the parametric form of the connectivity basis functions, where the scaling parameters $\boldsymbol{\theta}$ are unknown. Each connectivity basis function can, individually, be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get 

@@

To make the simplification, we first define

@@

Under the assumption that the basis function decomposition is accurate, the observation equation of the reduced model is found by substituting equation~\ref{DefFieldDecomp} into~\ref{eq:ObservationEquation} giving

@@

% \begin{equation}
% \mathbf{C} = \left[\begin{array}{{cccc}}
% \int_{\Omega}m(\mathbf{r}_1 - \mathbf{r}')\boldsymbol{\phi}_1(\mathbf{r}')d\mathbf{r}' & \int_{\Omega} m(\mathbf{r}_1 - \mathbf r')\boldsymbol \phi_2(\mathbf r')d\mathbf r' & \dots
% &\int_{\Omega}m(\mathbf r_1 - \mathbf r')\boldsymbol \phi_n(\mathbf
% r')d\mathbf r' \\
% \int_{\Omega}m(\mathbf r_2 - \mathbf r')\boldsymbol \phi_1(\mathbf
% r')d\mathbf r'&\int_{\Omega}m(\mathbf r_2 - \mathbf r')\boldsymbol
% \phi_2(\mathbf r')d\mathbf r'& \dots &\int_{\Omega}m(\mathbf r_2 -
% \mathbf r')\boldsymbol \phi_n(\mathbf r')d\mathbf r'\\
% \vdots&\vdots&\ddots&\vdots\\
% \int_{\Omega}m(\mathbf r_{n_y} - \mathbf r')\boldsymbol \phi_1(\mathbf
% r')d\mathbf r'&\int_{\Omega}m(\mathbf r_{n_y} - \mathbf r')\boldsymbol
% \phi_2(\mathbf r')d\mathbf r'& \dots &\int_{\Omega}m(\mathbf r_{n_y} -
% \mathbf r')\boldsymbol \phi_n(\mathbf r')d\mathbf r'
% \end{array}\right],
% \end{equation}
% and $m(\mathbf r_n - \mathbf{r}')$ is the observation kernel.
\section{Spectral Analysis and Model Selection}\label{SpectralAnalysisSection} Spectral analysis has been used to identify both the number of sensors and the number of basis functions required to reconstruct the neural field from sampled observations \cite{Sanner1992,Scerri2009}. Based on a two-dimensional extension of Shannon's sampling theorem \cite{Peterson1962}, the spatial bandwidth of the observed field can be used to provide a lower bound on both the number of sensors and the number of basis functions required to capture the dominant spectral characteristics of the neural field.

Let the spectral representation of the post-synaptic membrane voltage field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$. According to Shannon's sampling theorem the field must be spatially band-limited for an accurate reconstruction using spatially discrete observations. Such a condition ensures that $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$, where $\boldsymbol{\nu}_c$ is a cutoff frequency (typically taken as the -3~dB point) with $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c]^\top$. Given such a band-limited field, the distance between adjacent sensors, $\Delta_y$, must satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_y < \frac{1}{2\rho_y\nu_{c}}, 
\end{equation}
where $\rho_y \in \mathbb{R} \ge 1$ is an oversampling parameter. This condition must be satisfied to avoid spectral aliasing effects when reconstructing the hidden dynamic field $v_t(\mathbf{r})$ using the sampled observations, $\mathbf{y}_t$.

In practice, it is difficult to estimate the bandwidth of the cortex using traditional electrophysiological measurements, possibly preventing us from positioning sensors according to equation~\ref{eq:MinimumSensorDistance}. However, we envisage it may be possible to estimate the spatial bandwidth using other modalities with higher spatial resolution, such as fMRI, NIRS or other optical imaging techniques~\cite{Issa2000}. Nevertheless, spectral aliasing can still be avoided by a proper choice of the spatial sampling distance given the sensors' spectral characteristics. The spatial extent of the sensors results in spectral low-pass action, thus providing spatial anti-aliasing filtering. Such sensors therefore can only sense the field up to a known bandwidth, denoted by $\nu_{cy}$. Aliasing can then be avoided by positioning the sensors according to equation~\ref{eq:MinimumSensorDistance}, with $\nu_{cy}$ replacing $\nu_c$.

Although such sensors avoid errors due to aliasing, they attenuate the high spatial frequency variations in the observations. Therefore, any procedure applied to estimate the original field or the underlying connectivity structure from these band-limited observations has the potential to underestimate the high spatial frequency components of both the field and the connectivity kernel. This motivates the use of sensors with wider bandwidths (narrower in space). Nevertheless, this choice would require more sensors in order to satisfy Shannon's sampling theorem for a reasonable spatial region. Therefore, a compromise needs to be found between the bandwidth of the sensor response, the accuracy of the estimation results, the number of sensors used and the computational demands of the estimation procedure when designing experiments.

Similar considerations need to be made regarding the representation of the dynamic field, $v_t(\mathbf{r})$, using the basis function decomposition. Again applying Shannon's Sampling Theorem, the minimum distance between basis functions must satisfy 
\begin{equation}\label{eq:BasisFunctionSeparation}
	\Delta_{\phi} < \frac{1}{2\rho_{\phi}\nu_{cy}}
\end{equation}
where $\rho_{\phi} \in \mathbb{R} \ge 1$ is an oversampling parameter to determine the basis function separation.

The width parameter of the Gaussian basis functions can also be inferred using spectral considerations~\cite{Sanner1992,Scerri2009}. To demonstrate this, we begin with the two-dimensional Gaussian basis function defined as
\begin{equation}
 \phi(\mathbf r)=\mathrm{exp}\left({-\frac{1}{\sigma_{\phi}^2} \mathbf r^\top\mathbf r}\right)
\end{equation}
with the corresponding Fourier transform
\begin{equation}\label{eq:GaussianFT}
\boldsymbol\Phi(\boldsymbol \nu)=\left(\frac{1}{\pi\sigma_{\nu}^2}\right)\mathrm{exp}\left(-\frac{1}{\sigma_{\nu}^2}\boldsymbol\nu^\top \boldsymbol\nu\right),
\end{equation}
where 
\begin{equation}\label{eq:GaussianFTWidth}
	\sigma^2_{\nu} = \frac{1}{\pi^2\sigma_{\phi}^2}. 
\end{equation}
The basis function width, $\sigma^2_{\phi}$, should be chosen to give at least 3~dB attenuation at $\boldsymbol\nu_{cy}$ giving
\begin{equation}\label{eq:WidthCutOffRelationship}
 \sigma^2_{\phi}= \frac{1}{\pi^2\sigma_{\nu_{cy}}^2},
\end{equation}
where
\begin{equation}\label{eq:WidthFrequencyRelationship}
 \sigma^2_{\nu_{cy}}= \frac{2\boldsymbol\nu_{cy}^\top \boldsymbol\nu_{cy}}{\ln 2}.
\end{equation}
This ensures that the basis functions can represent the field with frequency content up to $\boldsymbol\nu_{cy}$. Derivations for equations~\ref{eq:GaussianFT} and \ref{eq:WidthFrequencyRelationship} are given in~\ref{ap:FrequencyAnalysis}.

Note that for spatially homogeneous isotropic fields, the basis functions can be placed on a regular grid. Thus, the knowledge of the distance between basis functions can be used to determine the total number of basis functions required to represent a known spatial region.

@@

The unscented Rauch Tung Striebel smoother (URTSS)~\cite{Sarkka2010} is used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF)~\cite{Julier1997, Merwe2003} in a forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward iteration to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and the second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{equation}\label{eq:sigmapoints1}
	\mathcal X_{0}=\bar x 
\end{equation}
\begin{equation}
	\mathcal X_{i}=\bar x+\left(\sqrt{( L + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, L 
\end{equation}
\begin{equation}\label{eq:sigmapoints2}
	\mathcal X_{i}=\bar x-\left(\sqrt{( L + \lambda)\mathbf P_x}\right)_{i- L}, \quad i= L+1, \dots, 2 L 
\end{equation}
where $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix from the filtering or smoothing, $\left(\sqrt{( L + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the weighted matrix square root of $\mathbf P_x$ and $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( L+\kappa) - L, 
\end{equation}
where the constant $\alpha$ determines the spread of the sigma points, $\beta$ incorporates prior knowledge of the distribution of $\mathbf{x}$ and $\kappa$ is a scaling parameter (see~\cite{Haykin2001} for more details).

@@

	\mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L, 

@@

   	\caption{\textbf{a.} Example of data generated by the full neural field model from the first 5 sensors. \textbf{b.} Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of intracranial EEG.} 

@@

\caption{Power in dB as a function of spatial frequency of the neural field. \textbf{a.} The average (over time) power of the spatial frequency of the neural field. \textbf{b.} The average (over time) power of the spatial frequency of the reconstructed neural field from the basis function decomposition. The reconstructed field shows a narrower bandwidth as a result of the band limiting properties of the sensors and basis functions.}

@@

One thing that is shown is that noise has a large impact in the estimation results. To reduce the effect of noise, an evoked response paradigm should be considered when using real data.

>>>>

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite dimensional state-space model. Spatial frequency analysis methods are introduced for model selection, by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation, and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic time constant with the proposed framework. The results also illustrate the loss of high spatial frequency information as the cost incurred by the model reduction procedure. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

Generating physiologically plausible neural field models are of great importance in order to study brain dynamics at the meso/macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's meso and macro-scale dynamics remains largely theoretical. Understanding the brain at this level is extremely important since it is this scale at which pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested.

@@

Unfortunately, the use of these models in the clinic has been limited, since they are constructed for ``general'' brain dynamics whereas pathologies almost always have unique underlying patient specific causes. Patient specific data from EEG is readily available in the clinical setting, suggesting an opportunity to make the patient-specific link to models of cortical dynamics. Furthermore, recent technological advances have driven an increased level of sophistication in recording techniques, with dramatic increases in spatial and temporal sampling~\cite{Brinkmann2009}. However the meso/macroscopic neural dynamic state is not directly observable in EEG data, making predictions of the underlying physiology inherently difficult.

For models to be clinically viable they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the Wilson and Cohen (WC)~\cite{Wilson1973} or Amari~\cite{Amari1977} models or a neural mass model like the Jansen and Ritt model~\cite{Jansen1995} to patient-specific EEG data. Fitting the neural models to individuals is a highly non-trivial task and until very recently, has not been reported in the literature.

@@

The paper proceeds by first deriving the continuum neural field equations in Section~\ref{NeuralModelSection}. Next, a finite dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. Section~\ref{SpectralAnalysisSection} establishes conditions using spatial frequency analysis for both sensor and basis function spacing and width, such that the dominant dynamics of the neural field can be represented by the reduced model. The state and parameter estimation procedure is described in Section~\ref{StateAndParameterEstimationSection}. The results for the spatial frequency analysis and parameter estimation are then presented in Section~\ref{ResultsSection}. The implications and limitations of this framework are discussed in Section~\ref{DiscussionSection} along with planned future developments.

@@

Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular due to being parsimonious yet having a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009,Qubbaj2007}. In certain regions of the cortex, each column is thought to be connected locally via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour~\cite{David2003,Bressloff2001,Breakspear2006}. Field models are also capable of generating complex patterns of activity such as Turing patterns, spirals, and travelling oscillations~\cite{Amari1977,Coombes2005,Coombes2007}.

@@

The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at position $\mathbf{r}$ to the local post synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations at locations $\mathbf{r}'$ can be described by 

@@

where $\zeta=\tau^{-1}$ and $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Nonlocal interactions between cortical populations are described by 

@@

where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel, and $\Omega$ is the spatial domain, representing a cortical sheet or surface. The connectivity kernel is typically a ``Mexican hat'' function which describes strong local activation, weak mid-range repression and weak long-range activation. The exact shape of this kernel is assumed to vary across patients, and hence needs to be inferred from data.

@@

   	\caption{A mexican-hat connectivity kernel, $w(\mathbf{r},\mathbf{r'})$ used in this paper. The function $w(.)$ is rotationally symmetric (isotropic) about zero, hence a cross section captures the important aspect of the connectivity kernel's shape. The kernel decays asymptotically to zero.} 

@@

To arrive at the final form of the model we shall express the synaptic response kernel as a Green's function 

@@

To arrive at the integro-difference equation (IDE) form of the model we discretise time using a first-order Euler method (see~\ref{Time Discretization}) giving 

@@

The mapping between the membrane voltage and the collected iEEG/LFP data is modelled using the observation function 
\begin{equation}
    \label{eq:ObservationEquation}
	\mathbf{y}_t =
	\int_{\Omega}{
	    m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)
	d\mathbf{r}'} + 
	\boldsymbol{\varepsilon}_t, 
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field, $n=1,...,N$ indexes the sensors, $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ and $\mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ denotes the multivariate normal distribution with mean zero and covariance matrix $\Sigma_{\varepsilon}$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the sensor pick-up geometry where 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}. 
\end{equation}

@@

Decomposition allows a continuous field to be represented by a finite dimensional state vector. This facilitates application of standard nonlinear state estimation methods such as the unscented Kalman filter. The field decomposition is described by 

@@

where $\boldsymbol{\psi}(r,r')$ is a vector of Gaussian basis functions and $\boldsymbol{\theta}$ is a vector of scaling parameters. By assuming a Gaussian isotropic connectivity structure the kernel basis functions can be written as $\psi(\mathbf{r}-\mathbf{r}')$. We will assume that we know the parametric form of the connectivity basis functions, where the scaling parameters $\boldsymbol{\theta}$ are unknown. Each connectivity basis function can, individually, be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get 

@@

To make the simplification we first define

@@

Using the same neural field basis function approximation, the observation equation of the reduced model is found by substituting equation~\ref{DefFieldDecomp} into~\ref{eq:ObservationEquation} giving

@@

% \begin{equation}
% \mathbf{C} = \left[\begin{array}{{cccc}}
% \int_{\Omega}m(\mathbf{r}_1 - \mathbf{r}')\boldsymbol{\phi}_1(\mathbf{r}')d\mathbf{r}' & \int_{\Omega} m(\mathbf{r}_1 - \mathbf r')\boldsymbol \phi_2(\mathbf r')d\mathbf r' & \dots
% &\int_{\Omega}m(\mathbf r_1 - \mathbf r')\boldsymbol \phi_n(\mathbf
% r')d\mathbf r' \\
% \int_{\Omega}m(\mathbf r_2 - \mathbf r')\boldsymbol \phi_1(\mathbf
% r')d\mathbf r'&\int_{\Omega}m(\mathbf r_2 - \mathbf r')\boldsymbol
% \phi_2(\mathbf r')d\mathbf r'& \dots &\int_{\Omega}m(\mathbf r_2 -
% \mathbf r')\boldsymbol \phi_n(\mathbf r')d\mathbf r'\\
% \vdots&\vdots&\ddots&\vdots\\
% \int_{\Omega}m(\mathbf r_{n_y} - \mathbf r')\boldsymbol \phi_1(\mathbf
% r')d\mathbf r'&\int_{\Omega}m(\mathbf r_{n_y} - \mathbf r')\boldsymbol
% \phi_2(\mathbf r')d\mathbf r'& \dots &\int_{\Omega}m(\mathbf r_{n_y} -
% \mathbf r')\boldsymbol \phi_n(\mathbf r')d\mathbf r'
% \end{array}\right],
% \end{equation}
% and $m(\mathbf r_n - \mathbf{r}')$ is the observation kernel.
\section{Spectral Analysis and Model Selection}\label{SpectralAnalysisSection} Spectral analysis was used to identify both the number of sensors and the number of basis functions required to reconstruct the neural field from sampled observations \cite{Sanner1992,Scerri2009}. Based on a two-dimensional extension of Shannon's sampling theorem \cite{Peterson1962}, the spatial bandwidth of the observed field can be used to provide a lower bound on both the number of sensors and the number of basis functions required to capture the dominant spectral characteristics of the neural field.

Let the spectral representation of the post-synaptic membrane voltage field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$. According to Shannon's sampling theorem the field must be spatially band-limited, for an accurate reconstruction using spatially discrete observations. Nevertheless, an approximate reconstruction can still be obtained if the field is approximately bandlimited with $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$, where $\boldsymbol{\nu}_c$ is a cutoff frequency (typically taken as the -3~dB point) with $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c]^\top$. Given such a band-limited field, the distance between adjacent sensors, $\Delta_y$, must satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_y < \frac{1}{2\rho_y\nu_{c}}, 
\end{equation}
where $\rho_y \in \mathbb{R} \ge 1$ is an oversampling parameter. This condition must be satisfied to avoid spectral aliasing effects when reconstructing the hidden dynamic field $v_t(\mathbf{r})$ using the sampled observations, $\mathbf{y}_t$.

In practice, it is difficult to estimate the bandwidth of the cortex using traditional electrophysiological measurements, possibly preventing us from positioning sensors in accordance to equation~\ref{eq:MinimumSensorDistance}. However, we envisage it may be possible to estimate it using other modalities with higher spatial resolution such as, fMRI, NIRS or other optical imaging techniques~\cite{Issa2000}. Nevertheless, spectral aliasing can still be avoided by a proper choice of the spatial sampling distance given the sensors spectral characteristics. The spatial response of all sensors result in a spectral low-pass action, thus providing spatial anti-aliasing filtering. Such sensors therefore bandlimit the field up to a known bandwidth, denoted by $\nu_{cy}$. Aliasing can then be avoided by positioning the sensors according to equation~\ref{eq:MinimumSensorDistance}, with $\nu_{cy}$ replacing $\nu_c$.

Although such sensors avoid errors due to aliasing, they attenuate the high spatial frequency variations in the observations. Therefore, any procedure applied to estimate the original field or the underlying connectivity structure from these band-limited observations has the potential to underestimate the high spatial frequency components of both the field and the connectivity kernel. Such a remark motivates the use of sensors with wider bandwidths (narrower in space). Nevertheless, this choice would require more sensors in order to satisfy Shannon's sampling theorem for a reasonable spatial region. Therefore, a compromise needs to be found between the bandwidth of the sensor response, the accuracy of the estimation results, the number of sensors used and the computational demands of the estimation procedure when designing experiments.

Similar considerations need to be made regarding the representation of the dynamic field, $v_t(\mathbf{r})$, using the basis function decomposition. Again applying Shannon's Sampling Theorem, the minimum distance between basis functions must satisfy 
\begin{equation}\label{eq:BasisFunctionSeparation}
	\Delta_{\phi} < \frac{1}{2\rho_{\phi}\nu_{cy}}
\end{equation}
where $\rho_{\phi}$ is an oversampling parameter to determine the basis function separation.

Such spectral considerations also facilitate the choice of the basis function width parameter~\cite{Sanner1992,Scerri2009}. To ensure that the basis functions can represent fields with frequency content up to $\boldsymbol\nu_{cy}$, they should be chosen to poses a spectral bandwidth of at least $\boldsymbol\nu_{cy}$. Such a condition is satisfied by setting the width parameter such that the spectral response of the basis functions results in an attenuation of not more than 3dBs at $\boldsymbol\nu_{cy}$.

To demonstrate this we begin with the 2-dimensional Gaussian basis function defined in equation \ref{eq:FieldBasisFunction}, with the corresponding Fourier transform
\begin{equation}\label{eq:GaussianFT}
\boldsymbol\Phi(\boldsymbol \nu)=\left(\frac{1}{\pi\sigma_{\nu}^2}\right)\mathrm{exp}\left(-\frac{1}{\sigma_{\nu}^2}\boldsymbol\nu^\top \boldsymbol\nu\right),
\end{equation}
where 
\begin{equation}\label{eq:GaussianFTWidth}
	\sigma^2_{\nu} = \frac{1}{\pi^2\sigma_{\phi}^2}. 
\end{equation}
Thus to obtain a 3dB attenuation at $\boldsymbol\nu_{cy}$, the basis function width, $\sigma^2_{\phi}$, should be chosen to be
\begin{equation}\label{eq:WidthCutOffRelationship}
 \sigma^2_{\phi}= \frac{1}{\pi^2\sigma_{\nu_{cy}}^2},
\end{equation}
where
\begin{equation}\label{eq:WidthFrequencyRelationship}
 \sigma^2_{\nu_{cy}}= \frac{2\boldsymbol\nu_{cy}^\top \boldsymbol\nu_{cy}}{\ln 2}.
\end{equation}
Derivations for equations~\ref{eq:GaussianFT} and \ref{eq:WidthFrequencyRelationship} are given in~\ref{ap:FrequencyAnalysis}.

Note that for a spatially homogeneous isotropic fields, the basis functions can be placed on a regular grid. Thus, the knowledge of the distance between basis functions directly implies the total number of basis functions required to represent a known spatial region.

@@

The unscented Rauch Tung Striebel smoother (URTSS)~\cite{Sarkka2010} is used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF)~\cite{Julier1997, Merwe2003} in a forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward iteration to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and the second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{equation}\label{eq:sigmapoints1}
	\mathcal X_{0}=\bar x 
\end{equation}
\begin{equation}
	\mathcal X_{i}=\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i, \quad i=1, \dots, L 
\end{equation}
\begin{equation}\label{eq:sigmapoints2}
	\mathcal X_{i}=\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L}, \quad i= L+1, \dots, 2 L 
\end{equation}
where $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix from the filtering or smoothing, $(\sqrt{( L + \lambda)\mathbf P_x})_i$ is the $i^{th}$ column of the weighted matrix square root of $\mathbf P_x$ and $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( L+\kappa) - L 
\end{equation}
where the constant $\alpha$ determines the spread of the sigma points, $\beta$ incorporates prior knowledge of the distribution of $\mathbf{x}$ and $\kappa$ is a scaling parameter (see~\cite{Haykin2001} for more details).

@@

	\mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L. 

@@

   	\caption{\textbf{a.} Example of data generated by the full neural field model from the first 5 sensors. \textbf{b.} Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of iEEG.} 

@@

\caption{Spatial frequency of the neural fields. \textbf{a.} The average (over time) spatial frequency of the neural field. \textbf{b.} The average (over time) spatial frequency of the estimated neural field. As a result of the band limiting properties of the sensors and basis functions the reconstructed field shows a narrower bandwidth.}

@@
